Phenotype and risk assesment of insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry

### 2019-09-09 analysis/paper/paper.Rmd
V2.1 - gramarly corrected

---
title: "Phenotypes and risk assesment of insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik^1^
  - Sabine Dölle-Bierke^1^
  - Franziska Ruëff^2^ 
  - Claudia Pföhler^3^
  - Kathrin Scherer Hofmeier^4^
  - Margitta Worm^1^
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    #bookdown::html_document2:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/expert-review-of-clinical-immunology.csl" # Insert path for the bib-style
---

@@

Hypersensitivity to insect venom can manifest as a systemic reaction (anaphylaxis) in up to 0.3-7.5% of the adult population [@Bil2008]. Insect venom anaphylaxis (IVA) can be fatal, and patients sometimes require lifelong systemic immunotherapy [@Sturm2017].

Recent expert position on the future of anaphylaxis research indicates the need for a more precise description of the diagnosis, biomarkers, and phenotypes of anaphylaxis [@JimenezRodriguez2018]. Nevertheless, in order to facilitate a precision-medicine approach [@Muraro2017] in the treatment of anaphylaxis - it is first needed to study its clinical subtypes and pathomechanisms in more detail.

Clinical controlled trials in anaphylaxis are hardly possible due to the acuteness of this life-threatening condition. Therefore large registries, gathering clinical data as well as biological samples from patients with a well documented (recent) history of anaphylaxis are a crucial tool in investigating this entity.

@@

We searched the European Anaphylaxis Registry [@Grabenhenrich2016] database from March 2019 for anaphylaxis cases elicited by insect's venom. The flowchart in figure \@ref(fig:flow) represents the detailed case-selection process.

The final database consisted of `r countries %>% summarise(sum(n))` cases of insect elicited anaphylaxis from `r countries %>% summarize(length(n))` countries. Severe reactions were identified based on the definition by NIAID/FAAN [@Sampson2006] or as Ring and Messmer Scale - grades III and IV) and presented with significant hypoxia, hypotension, confusion, collapse and loss of consciousness, or incontinence.

The cases were either assigned to IVA group (if the triggering factor was an insect sting) or non-IVA group (other elicitors of anaphylaxis).
We compared the frequency of various symptoms, cofactors --- known to increase the risk of severe anaphylaxis, [@Worm2018] and management in both groups.

Due to a large number of documented reactions in the European Anaphylaxis registry - we were able to match the IVA cases to selected non-IVA cases according to sex and age to reduce the comparison bias. Subsequently, we compared the management in both groups and matched the control group according to the severity of a reaction. Propensity score matching has been performed using the "MatchIt" package for R [@matchit]. The results of the propensity score matching are in Fig. \@ref(fig:flow)B-D.

<!-- ## Data handling and statistics -->
We used the R Statistical Package [@R] for statistical analysis. A simple comparison of categorical variables was performed using either Chi^2^ test or Fisher's exact test (where the number of observation in a bin was less than 10), continuous variables were analyzed using Mann-Whitney U test. We defined statistical significance as α = 0.05. Data, along with the analysis script, can be accessed online at https://github.com/wolass/venomanaphylaxiscompendium.

@@

IVA was more frequent in adults and seniors than children and young adults. (Fig \@ref(fig:insectstime)B). When we analyzed a density plot of IVA cases according to age - we determined a bimodal distribution forming two subsets of patients with a cutoff age of 22 (Fig \@ref(fig:flow)B.

@@

Patients with concomitant mastocytosis had significantly more often anaphylaxis without skin symptoms (namely urticaria and flushing) when compared to patients without diagnosed mastocytosis
(`r (skin_symptoms$masto_tab$prop[c(2,1)]*100) %>% round(1) %>% paste0(collapse="% vs. ")`%, `r pval(skin_symptoms$masto_chisq$p.value)`). We decided to further investigate cases where mastocytosis was not diagnosed. In these patients urticaria or flushing as symptoms of anaphylaxis were present more often in IVA than anaphylaxis triggered by other elicitors (`r (skin_symptoms$tab$prop[2]*100) %>% round(1)`% vs `r (skin_symptoms$tab$prop[4]*100) %>% round(1)`% respectively, `r pval(skin_symptoms$chisq$p.value)`).  Morover, in patients without concomitant mastocytosis who were also lacking skin symptoms, anaphylaxis was significantly more frequently severe (`r (skin_symptoms$severe_tab$prop[c(2,6)]*100) %>% round(1) %>% paste0(collapse="% in IVA vs. ")`%, `r pval(skin_symptoms$masto_chisq$p.value)`% in non-IVA). We performed a factorial logistic regression modelling and confirmed the significant interaction effect between the presence of skin symptoms and the trigger (eiter IVA or non-IVA) on the severity of anaphylaxis (`r pval(skin_symptoms$severity_fit_skin_symptoms %>% coef %>% .[4,4])`). Therefore, non-mastocytic patients presenting without urticara or flushing tended to have more severe anaphylaxis when it was triggered by insects, but not other elicitors of anaphylaxis (Fig. \@ref(fig:figsymptoms)C).

Following this observation, we decided to investigate the association of skin symptoms with the tryptase levels in patients who did not have a diagnosis of mastocytosis. For this comparison, we removed the cases with BST above 11.5 ng/ml, potentially indicating non-diagnosed mastocytosis. Similarly, tryptase levels correlated with the severity of anaphylaxis, were higher in IVA patients, and the effect was prominently visible for IVA patients (`r skin_symptoms$fit_tryp[[1]][4,5] %>% pval()`). We did not observe such an interaction in the non-IVA group.

@@

The factor most prominently associated with an increased risk of severe anaphylaxis was mastocytosis, and there were no differences in elicitor groups (Fig. \@ref(fig:figcofactors)). However, mastocytosis increased the risk of cardiac arrest in patients undergoing IVA significantly more than in patients undergoing anaphylaxis due to other triggers (Fig. S\@ref(fig:supplement)).

Concomitant diseases were more prevalent in IVA than in non-IVA cases (`r age_sex_matched %>% count(grouping,q_410_cardio_cur) %>% group_by(grouping) %>% mutate(prop= n/sum(n)) %>% {.[c(2,5),4]} %>% pull() %>% compare()`) and were associated with a higher risk of severe anaphylaxis when elicited by insects but were irrelevant in non-IVA cases (Fig. \@ref(fig:figcofactors)). Interestingly, BST values were increased in patients with concomitant cardiovasular diseases, irrespectively of the reaction severity (Fig. S\@ref(trypcardiocur)).

The risk of severe anaphylaxis in patients concomitantly using ACE-I (as well as beta-blockers) could not be independently measured due to coexisting cardiovascular pathologies. ACE-I use was, however, more often associated with cardiac arrests in all anaphylaxis cases (`r ace_tab %>% filter(q_114_cardiac_arrest =="yes") %>% dplyr::select(prop) %>% pull() %>% compare()`, `r pval(ace_tab %>% dplyr::select(-prop) %>% spread(q_114_cardiac_arrest,n) %>% {chisq.test(.[,2:3])$p.value})`) and there were no differences between IVA nad non-IVA  (Fig. \@ref(fig:figcofactors)C). 
Beta-blocker use was generally associated with higher severity of anaphylaxis and with the onset of cardiovascular symptoms (cardiac arrest, chest pain), but we saw no difference between groups `r summary(beta_fit)[["coefficients"]][3,4] %>% pval()`). Surprisingly, arrhythmia was more frequently reported in patients undergoing IVA with concomitant beta-blockers (Fig. \@ref(fig:figcofactors)B).

Baseline serum tryptase (BST) correlated with the severity of reactions (Ring and Messmer scale). Moreover, elevated BST was more prominently increasing the risk of severe anaphylaxis in IVA patients than in other cases (Fig. \@ref(fig:figcofactors)C).

Cases with cardiac arrest were associated with an increased tryptase above 8 ng/ml, and this proportion was higher in IVA when compared to other elicitors. Loss of consciousness was a symptom associated with increased tryptase levels but only in patients with IVA (Fig. \@ref(fig:figcofactors)C). Based on the severity and symptom profile, we decided to use a tryptase cut off value of 8 ng/ml (Fig. \@ref(fig:figcofactors)B) rather than the currently used 11.5 ng/ml.

@@

<!-- Symptoms -->
<!-- Cardio symptoms phenomenon -->
IVA was more often associated with cardiologic symptoms non-IVA.
Previous studies suggest [@Worm2018; @nittner2015insect; @Bonadonna2010] an essential link between the cardiovascular system and insect sting hypersensitivity. 
IVA has been associated with cardiac arrhythmias usually occurring in patients with preexisting heart disease [@Sharma2016] and Kounis syndrome (coronary arterial spasm induced by the release of mast cell mediators [@CDT1609; @sinkiewicz2008allergic]).

@@

Importantly, the absence of skin symptoms was associated with more severe IVA, which was also after we excluded patients with a known diagnosis of mastocytosis. Subsequently, we correlated the severity of anaphylaxis with BST levels and identified an interaction between the absence of skin symptoms and IVA using generalized linear regression.

@@

Regarding the factors increasing risk of IVA - older patients experienced IVA more frequently. Young patients mainly suffer from food-induced anaphylaxis [@Grabenhenrich2016]. One of the limitations of our registry is the fact that we can only compare cases of anaphylaxis and due to the lack of healthy cohort - cannot draw conclusions on the frequency of a particular type of anaphylaxis in the population.

@@

When evaluating the risk of future severe episodes - baseline serum tryptase levels over ng/ml should be considered.

@@

@@

```{r figsymptoms, fig.cap="Symptoms of insect venom anaphylaxis (IVA) compared to other elicitors. A: Proportional presentation of specific reaction symptoms. B: High-level overview of involved organ systems and selected cofactors in the form of a radar plot.", fig.width=8.5, fig.height = 6.2}
  fig_symptoms2
```

```{r figcofactors, fig.cap="Cofactors of insect venom anaphylaxis. A: Odds ratios of eliciting severe anaphylaxis. B: Proportion of cases elicited by insects or other elicitors (upper panels) according to tryptase levels and cardiologic symptoms. ", fig.width = 6.7, fig.height = 7.4}

@@

```{r supplement, fig.width= 6, fig.height= 5, fig.cap="Symptoms of anaphylaxis"}

@@

```{r rfplot}
rf_ggplot_treatment
```

```{r}
adren_prev
```

>>>>

---
title: "Phenotype and risk assesment of insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik^1^
  - Sabine Dölle-Bierke^1^
  - Franziska Ruëff^2^ 
  - Claudia Pföhler^3^
  - Kathrin Scherer Hofmeier^4^
  - Margitta Worm^1^
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    #bookdown::html_document2:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/expert-review-of-clinical-immunology.csl" # Insert path for the bib-style
---

@@

Hypersensitivity to insect venom can manifest as a systemic reaction (anaphylaxis) in up to 0.3--7.5% of the adult population [@Bil2008]. Insect venom anaphylaxis (IVA) can be fatal, and patients sometimes require lifelong systemic immunotherapy [@Sturm2017].

Recent expert position on the future of anaphylaxis research indicates the need for a more precise description of the diagnosis, biomarkers, and phenotypes of anaphylaxis [@JimenezRodriguez2018]. Nevertheless, in order to facilitate a precision-medicine approach [@Muraro2017] in the treatment of anaphylaxis --- it is first needed to study its clinical subtypes and pathomechanisms in more detail.

Controlled clinical trials in anaphylaxis are hardly possible due to the acuteness of this life-threatening condition. Therefore large registries, gathering clinical data as well as biological samples from patients with a well documented (recent) history of anaphylaxis are a crucial tool in investigating this entity.

@@

We searched the European Anaphylaxis Registry [@Grabenhenrich2016] database from March 2019 for anaphylaxis cases elicited by insect's venom. The flowchart in Fig. \@ref(fig:flow) represents the detailed case-selection process.

The final database consisted of `r countries %>% summarise(sum(n))` cases of insect-elicited anaphylaxis reported from allergology centers in `r countries %>% summarize(length(n))` countries. We based the diagnosis of anaphylaxis on the definition by NIAID/FAAN [@Sampson2006] and the severity according to Ring and Messmer Scale [@Ring1977]. Grades III and IV (presenting with significant hypoxia, hypotension, confusion, collapse and loss of consciousness, or incontinence or cardiac arrest) were considered severe.

The cases were either assigned to IVA group (if the triggering factor was an insect sting) or non-IVA group (if other elicitors triggered anaphylaxis).
We compared the frequency of various symptoms, cofactors --- known to increase the risk of severe anaphylaxis, [@Worm2018] and management in both groups.

Due to a large number of documented reactions in the European Anaphylaxis registry - we were able to match the IVA cases to selected non-IVA cases according to sex and age to reduce the comparison bias. Subsequently, we compared the management in both groups and matched the control group according to the severity of a reaction. Propensity score matching has been performed using the "MatchIt" package for R [@matchit]. The results of the propensity score matching are in Fig. \@ref(fig:flow)B-D and supplementary Fig \@ref(fig:matchingresult).

<!-- ## Data handling and statistics -->
We used the R Statistical Package [@R] for statistical analysis. A simple comparison of categorical variables was performed using either Chi^2^ test or Fisher's exact test (where the number of observation in a bin was less than 10). Continuous variables were analyzed using Mann-Whitney U test. In case of comparisons with two or more independent variables, we used Factorial ANOVA or Generalized Linear Models. We defined statistical significance as α = 0.05. Data, along with the analysis script, can be accessed online at https://github.com/wolass/venomanaphylaxiscompendium.

@@

IVA was more frequent in adults and seniors than children and young adults. (Fig \@ref(fig:insectstime)B). When we analyzed a density plot of IVA cases according to age - we determined a bimodal distribution forming two subsets of patients with a cutoff age of 22 (Fig \@ref(fig:flow)B).

@@

Patients with concomitant mastocytosis had significantly more often anaphylaxis without skin symptoms (namely urticaria and flushing) when compared to patients without diagnosed mastocytosis (`r (skin_symptoms$masto_tab$prop[c(2,1)]*100) %>% round(1) %>% paste0(collapse="% vs. ")`%, `r pval(skin_symptoms$masto_chisq$p.value)`). We decided to further investigate non-mastocytic patients. In these patients urticaria or flushing as symptoms of anaphylaxis were present less often in IVA than anaphylaxis triggered by other elicitors (`r (skin_symptoms$tab$prop[2]*100) %>% round(1)`% vs `r (skin_symptoms$tab$prop[4]*100) %>% round(1)`% respectively, `r pval(skin_symptoms$chisq$p.value)`).  Morover, in patients without concomitant mastocytosis who were also lacking skin symptoms, anaphylaxis was significantly more frequently severe (`r (skin_symptoms$severe_tab$prop[c(2,6)]*100) %>% round(1) %>% paste0(collapse="% in IVA vs. ")`%, `r pval(skin_symptoms$masto_chisq$p.value)`% in non-IVA). We performed a factorial logistic regression modelling and confirmed the significant interaction effect between the presence of skin symptoms and the trigger (eiter IVA or non-IVA) on the severity of anaphylaxis (`r pval(skin_symptoms$severity_fit_skin_symptoms %>% coef %>% .[4,4])`). Therefore, non-mastocytic patients presenting without urticara or flushing tended to have more severe anaphylaxis when it was triggered by insects, but not other elicitors of anaphylaxis (Fig. \@ref(fig:figsymptoms)C).

Following this observation, we decided to investigate the association of skin symptoms with the tryptase levels in patients who did not have a diagnosis of mastocytosis. For this comparison, we removed the cases with BST above 11.5 ng/ml, potentially indicating non-diagnosed mastocytosis. Similarly, tryptase levels correlated with the severity of anaphylaxis, were higher in IVA patients, and the effect was prominently visible for IVA patients (`r skin_symptoms$fit_tryp[[1]][4,5] %>% pval()`). We did not observe such an interaction in the non-IVA group (Fig. \@ref(fig:figsymptoms)D).

@@

The factor most prominently associated with an increased risk of severe anaphylaxis was mastocytosis, and there were no differences in elicitor groups (Fig. \@ref(fig:figcofactors)). However, mastocytosis increased the risk of cardiac arrest in patients undergoing IVA significantly more than in patients undergoing anaphylaxis due to other triggers (Fig. S\@ref(fig:supplement)A).

Concomitant cardiovascular diseases were more prevalent in IVA than in non-IVA cases (`r age_sex_matched %>% count(grouping,q_410_cardio_cur) %>% group_by(grouping) %>% mutate(prop= n/sum(n)) %>% {.[c(2,5),4]} %>% pull() %>% compare()`) and were associated with a higher risk of severe anaphylaxis when elicited by insects but were irrelevant in non-IVA cases (Fig. \@ref(fig:figcofactors)). Interestingly, BST values were increased in patients with concomitant cardiovasular diseases, irrespectively of the reaction severity (Fig. S\@ref(fig:trypcardiocur)).

The risk of severe anaphylaxis in patients concomitantly using ACE-I (as well as beta-blockers) could not be independently measured due to coexisting cardiovascular pathologies. ACE-I use was, however, more often associated with cardiac arrests in all anaphylaxis cases (`r ace_tab %>% filter(q_114_cardiac_arrest =="yes") %>% dplyr::select(prop) %>% pull() %>% rev() %>% compare()`, `r pval(ace_tab %>% dplyr::select(-prop) %>% spread(q_114_cardiac_arrest,n) %>% {chisq.test(.[,2:3])$p.value})`) and there were no differences between IVA nad non-IVA  (Fig. \@ref(fig:figcofactors)C). 
Beta-blocker use was generally associated with higher severity of anaphylaxis and with the onset of cardiovascular symptoms (cardiac arrest, chest pain), but we saw no difference between groups `r summary(beta_fit)[["coefficients"]][3,4] %>% pval()`). Surprisingly, arrhythmia was more frequently reported in patients undergoing IVA with concomitant beta-blockers (Fig. \@ref(fig:figcofactors)C).

Baseline serum tryptase (BST) correlated with the severity of reactions (Ring and Messmer scale). Moreover, elevated BST was more prominently increasing the risk of severe anaphylaxis in IVA than in non-IVA (Fig. \@ref(fig:figsymptoms)D and Fig. \@ref(fig:figcofactors)B).
Cases with cardiac arrest were associated with an increased tryptase above 8 ng/ml, and this proportion was higher in IVA when compared to other elicitors. Loss of consciousness was a symptom associated with increased tryptase levels but only in patients with IVA (Fig. \@ref(fig:figcofactors)C). Based on the severity and symptom profile, we decided to use a tryptase cut off value of 8 ng/ml (Fig. \@ref(fig:figcofactors)B) rather than the currently used 11.5 ng/ml.

@@

<!-- Symptoms -->
<!-- Cardio symptoms phenomenon -->
IVA was more often associated with cardiologic symptoms non-IVA.
Previous studies suggest [@Worm2018; @nittner2015insect; @Bonadonna2010] an essential link between the cardiovascular system and insect sting hypersensitivity. 
IVA has been associated with Kounis syndrome (coronary arterial spasm induced by the release of mast cell mediators [@CDT1609; @sinkiewicz2008allergic]) and cardiac arrhythmias usually occurring in patients with preexisting heart disease [@Sharma2016].

@@

Importantly, the absence of skin symptoms was associated with more severe IVA, which was also true after we excluded patients with a known diagnosis of mastocytosis. Subsequently, we correlated the severity of anaphylaxis with BST levels and identified an interaction between the absence of skin symptoms and IVA using generalized linear regression.

@@

Regarding the factors increasing risk of IVA - older patients experienced IVA more frequently. Young patients mainly suffer from food-induced anaphylaxis [@Grabenhenrich2016]. One of the limitations of our registry is the fact that we can only compare cases of anaphylaxis and due to the lack of healthy cohort --- cannot draw conclusions on the frequency of a particular type of anaphylaxis in the population.

@@

When evaluating the risk of future severe episodes - patients with BST over 8 ng/ml should undergo extensive diagnostic tests to exclude ISM or MCAS and should be provided with two adrenaline autoinjectors.

@@

```{r figsymptoms, fig.cap="Symptoms of insect venom anaphylaxis (IVA) compared to other elicitors. A: Proportional presentation of specific reaction symptoms. B: High-level overview of involved organ systems and selected cofactors in the form of a radar plot. C: Interaction effect of IVA and lack of skin symptoms (urticaria or flushing) on the severity of anaphylaxis. D: Baseline Serum Tryptase (BST) increases with the severity of anaphylaxis a lot more prominently in IVA.", fig.width=8.5, fig.height = 6.2}
fig_symptoms2
```

@@

```{r figcofactors, fig.cap="Cofactors of insect venom anaphylaxis. A: Odds ratios of eliciting severe anaphylaxis. B: Proportion of cases elicited by insects or other elicitors (upper panels) according to tryptase levels and cardiologic symptoms. ", fig.width = 7, fig.height = 8}

@@

```{r matchingresult, fig.cap="Results of matching the cohort according to sex and age in order to perform a case-controlled study. A: The original distribution of IVA and non-IVA cases according to age group and sex. B: The distribution of IVA and non-IVA after age and sex matching with the use of MatchIt package for R."}
matching.result 
```

```{r supplement, fig.width= 6, fig.height= 5, fig.cap="Symptoms of anaphylaxis. A: The association between cardiac arrest and concomitant mastocytosis in IVA and non-IVA. B: Hypotension frequency in two age groups of anaphylaxis. C: Crammer's V as the measure of association between groups anaphylaxis (IVA vs. non-IVA)."}

@@

```{r rfplot, fig.cap="Therapy of anaphylaxis. A: Adrenaline use considering if patients had previous anaphylaxis. B: Variable impo"}
ggarrange(
  adren_prev+
    geom_segment(aes(x = 0.5,
                     xend = 2.5,
                     y = 0.33,
                     yend = 0.33),
                 color="white",
                 linetype =2),
  rf_ggplot_treatment,
  nrow = 1,
  ncol = 2,
  labels=c("A","B")
)
```
